---
title: 'POLI 178: Afghanistan Military Spending EDA + Statistical Inference'
author: "Baraa Zekeria"
date: "December 2, 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Afghanistan: Military Expenditure as Share of GDP

  - The effect of US intervention on Afghanistan's military spending
  - Military spending is intimately linked to war. In times of peace, few countries spend more than 5% of
  GDP on their military, choosing instead to maintain their capabilities and make use of their armies in 
  other ways.

 [Our World Data - Violence and War](https://ourworldindata.org/military-spending)
 
# Clear Environment

```{r}
rm(list=ls(all=TRUE))
```

# Load Package 

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(sjPlot)
library(vtable)
library(broom)
library(stargazer)
```

# Load Data

```{r message=FALSE, warning=FALSE}
mil_exp <- read_csv("military-expenditure-as-share-of-gdp.csv")
```

# View Data

```{r}
glimpse(mil_exp)
```

```{r}
colnames(mil_exp)
```

# Data Cleaning

# Change the **class()** of the ```Year``` and ```pct_gdp``` columns

```{r}
mil_exp$Year <- as.numeric(mil_exp$Year)
mil_exp$pct_gdp <- as.numeric(mil_exp$pct_gdp)
```

## Confirm the data type changes

```{r}
class(mil_exp$Year)

class(mil_exp$pct_gdp)
```



## Drop the ```code``` column given its redundance for this analysis

```{r}
mil_exp <- subset(mil_exp, select = -c(Code))
```

### Confirm the drop by calling the **colnames()** function

```{r}
colnames(mil_exp)
```


## Subset the DF to only include where the ```Entity``` is ```Afghanistan```  and re-assign it to the ```mil_exp``` object

```{r}
mil_exp <- mil_exp %>%
  filter(Entity == "Afghanistan")
```

## Create a new column, ```intervention_type```, given the ```Year``` values

```{r}
mil_exp$intervention_type <- ifelse(mil_exp$Year >= 2001, "Post-US Intervention", "Pre-US Intervention")
```


# Summary Statistics 

## Since the focus of this project is related to US intervention of analysis starting in 2001, find the diifference in means for ```pct_GDP```

```{r}
# Filter to 2 separate DF's
pre_us <- mil_exp %>%
  filter(Year < 2001) 

post_us <- mil_exp %>%
  filter(Year >= 2001) 
```

```{r warning=FALSE}
# Calculate the means
avg_pct_gdp <- mean(mil_exp$pct_gdp)

pre_us_avg_pct_gdp <- mean(pre_us$pct_gdp)

post_us_avg_pct_gdp <- mean(post_us$pct_gdp)

```

  - Means
    - Overall: `r avg_pct_gdp`%
    - Pre-US (< 2001): `r pre_us_avg_pct_gdp`%
    - Post-US (>= 2001): `r post_us_avg_pct_gdp`%
    - Difference: `r pre_us_avg_pct_gdp - post_us_avg_pct_gdp`%

```{r}
st(mil_exp, vars = c("Year", "pct_gdp"))
```


# Visualization

```{r}
mean_gdp <- mil_exp %>% 
  group_by(intervention_type) %>%
  summarise(mean_gdp_rate = mean(pct_gdp, na.rm = TRUE))

mean_gdp
```


```{r}
mil_exp %>% 
  group_by(intervention_type) %>%
  summarise(mean_gdp_rate = mean(pct_gdp, na.rm = TRUE)) %>%
  ggplot(aes(intervention_type, mean_gdp_rate, fill = intervention_type)) +
  geom_bar(stat ="identity") + 
  scale_x_discrete(limits = c("Pre-US Intervention", "Post-US Intervention")) +
  labs(title = "Average Afghanistan Military Expenditure (% of GDP) by Intervention
      Type", x = "Year", y = "Average Military Expenditure (% of GDP)",
      caption = "Source: Our World in Data")
```

The average military expenditure for *Pre-US Intervention* is 1.81% and *Post-US Intervention* is 1.61%.

```{r}
ggplot(data = mil_exp, 
       aes(x = Year, 
           y = pct_gdp, 
           color = intervention_type)) +
  geom_point() +
  geom_smooth(method = lm, se = F) + 
  geom_line() + 
  labs(title = "Afghanistan Military Expenditure (% of GDP) by Year",
       x = "Year",
       y = "Military Expenditure (% of GDP)",
       caption = "Source: Our World in Data")
```

Before the US intervention on Afghanistan, we notice a roughly positive correlation in military spending from 1970-1977. The, from 2004-2009, during that 6 year period, we see an up and down trend with military spending. Though the data changes, come 2010, where military spending continuously decreases from ~ 1.9% - 1.1 %. We see a slight increase in 2014 and 2018, 1.08% to 1.30% and 0.95% to 1.01%, respectively.

*Note: Data was not taken into account from 1978-2003*

# Statistical Inference

```{r}
# Save the GDP values in an array
pre_us_vals <- pre_us$pct_gdp

post_us_vals <- post_us$pct_gdp
```

```{r}
# T-Test Difference in Means
tt <- t.test(as.numeric(pre_us_vals), as.numeric(post_us_vals))

tidy(tt)
```

```{r}
# Overall
lm_overall <- lm(pct_gdp ~ Year, data = mil_exp)

lm_overall
```

```{r message=FALSE, warning=FALSE}
# Pre-US Intervention
lm_pre <- lm(filter(mil_exp, intervention_type == "Pre-US Intervention")$pct_gdp ~ filter(mil_exp, intervention_type == "Pre-US Intervention")$pct_gdp)

lm_pre
```

```{r message=FALSE, warning=FALSE}
# Post-US Intervention
lm_post <- lm(filter(mil_exp, intervention_type == "Post-US Intervention")$pct_gdp ~ filter(mil_exp, intervention_type == "Post-US Intervention")$pct_gdp)

lm_post
```

```{r message=FALSE, warning=FALSE}
# Combine the linear regression models
stargazer(lm_overall, lm_pre, lm_post, type = "html", dep.var.labels = c("Overall", "Pre-US Intervention", "Post- US Intervention"), title = "Linear Regressions", out = "mil_regressions.htm")
```

